Local Spectral Clustering of Density Upper Level Sets

### 2019-01-23 icml/icml_local_density_clustering.tex
final icml version

\begin{abstract}
\vskip 0.1 in % RJT: there was no space here and it looked very squished 
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.
\end{abstract}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be been computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

@@

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citet{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two critera capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the rotation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationariy distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ t: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}},  
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

@@

For a level $\beta > 0$ and a target volume $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\widehat{C} = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

\item In Section \ref{section: examples}, we discuss the implications of our
  results for some example density functions of interest, and empirically
  demonstrate that violations of the gometric conditions we require manifestly
  impact density cluster recovery. 
\end{enumerate}

@@

\subsection{Related Work}

@@

\citet{vonluxburg2008,hein2005}, building on earlier work of
\citet{koltchinskii2000}, studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

@@

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exist $\gamma, c_1 > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq c_1 \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.
% RJT: should we change this to c_0?  c_1 seems to be used below and throughout
% for other constants, and c_0 is not used

@@

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} and \ref{asmp:  
cluster_separation} combined are similar to the $(\sigma,\epsilon)$-saliency of
\citet{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citet{singh2009} (as well as
many other works on density clustering and level set estimation.) It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as PPR.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on (thickened) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will abbreviate $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and similarly, $\Psi(S; G_{n,r})$ by $\Psi_{n,r}(S)$, and
$\tau_{\infty}(G_{n,r}[S])$ by $\tau_{n,r}(S)$. We will also use $\nu$ for
Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for the measure of the unit ball
$B=B(0,1)$.  
% RJT: you had B(0,d) before ... just check this should have been B(0,1)

@@

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_1, \gamma > 0$. Then for any $0 < r <
\sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}
  { \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
we have
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$, where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

@@

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is almost
tight. Specifically, choosing $\Asig = B(0,\sigma)$ and
$$
f(x) = \lambda - \dist(x,\Asig)^{\gamma},
$$
% RJT: is this correct (is this what you meant)?  Original formula is below
% $$
% f(x)  = \begin{cases}
%   \lambda &\text{for $x \in \Asig$}, \\
%   \lambda - \dist(x,\Asig)^{\gamma} & \text{ for $0 < \dist(x,\Asig) < r$}, 
% \end{cases}
% $$
we have that for $n$ on the order of the lower bound in \eqref{eqn: 
  conductance_sample_complexity}, 
$$
  \frac{\Phi_{n,r}(\Asig[\mathbf{X}])}{r} \geq c \frac{(\lambda -
    \frac{r^{\epsilon+1}}{\epsilon+1})}{\lambda} - \epsilon,
$$
with probability at least $1 - \delta$, for some constant $c$. (Note that the
factor of $1 / \sigma$ in $c_{\sigma}$ is not replicated above.)
\end{remark}

@@

Next we lower bound the inverse mixing time \eqref{eqn: inv_mixing_time} of 
$\Csig[\Xbf]$, or equivalently, as $\Psi_{n,r}(\Csig[\Xbf]) =
1/\tau_{n,r}(\Csig[\Xbf])$, we upper bound the mixing time. 
% RJT: note I changed the next two theorems to be upper bounds on the mixing
% time.  Upper bounds on the inverse inverse mixing time, which was a bit
% superfluous.  Just make sure that the supplement is in sync with this

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound_nonconvex}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D > 0$. Then for any $0 <
r < \sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and $n$ satisfying 
$$
\sqrt{3^{d+1}\frac{(\log n + d\log \mu + \log(4/\delta))}{n \nu_d r^d
    \lambda_{\sigma}}} \leq \epsilon, 
$$
where \smash{$\mu = \log(\frac{2D}{r})$}, we have
\begin{multline}
\label{eqn: inverse_mixing_time_lower_bound_nonconvex}
\tau_{n,r}(\Csig[\Xbf]) \leq (d \log \mu + c_{\lambda}) \cdot \\  
\left(c_1 + c_2 \frac{\Lambda_{\sigma}^4D^{2d}}{\lambda_{\sigma}^4r^{2d}}\left(d
    \log \mu + c_{\lambda}\right)\right),
\end{multline}
with probability at least $1 - \delta$, where $c_1,c_2>0$ are constants
depending only on the dimension $d$, and $c_{\lambda} = \log(\Lambda_{\sigma}^2/ 
\lambda_{\sigma}^2)$.   
\end{theorem}
% RJT: saying that c_1,c_2 depend on the dimension is a little odd/weak.
% c_2 multiplyies, e.g., a factor of d \log \mu.  So if it's d^{10}, then this
% will swamp the linear factor of ....

\begin{remark}
The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} relies
on upper bounding the mixing time using the \emph{conductance} of $G_{n,r}[\Csig[X]]$, 
$$
\widetilde{\Phi} = \min_{S \subseteq \Csig[\Xbf]} \Phi(S; G_{n,r}[\Csig[X]]).
$$
The factor of $1/r^{2d}$ in the bound in \eqref{eqn:
  inverse_mixing_time_lower_bound_nonconvex} is suboptimal. This exponential 
dependence on $d$ stems from a loose bound on the aforementioned
conductance. In particular, we assert only that any set $S \subseteq \Csig[X]$
must have $\cut(S; \Csig[\Xbf])$ on the order of $n^2 r^{2d}$, while upper bounding
$\vol(S; \Csig[\Xbf])$ by roughly $n^2 r^d$, for a bound on the conductance of
order $r^d$. The presence of $r^{2d}$ comes from upper bounding the mixing time
by about \smash{$1/\widetilde{\Phi}^2$}, this being a variant of classic results
on rapid mixing \citep{jerrum89}.  
\end{remark}

@@

\begin{remark}
The only potentially exponential dependence on dimension comes from the
factor of $c_d$. However, for sufficiently small values of $r$ this will
be dominated by the preceding factors, due to the presence of the
$o_r(1)$ term. 
\end{remark}
% RJT: Don't get this comment.  You have o_r(1) / r^2, so this might still blow
% up as r -> 0, right?

\begin{remark}
We achieve superior rates in Theorem \ref{thm: inverse_mixing_time_lower_bound}
to those in Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} in part 
by working with a generalization of the conductance, the \emph{conductance
  function},
$$
\widetilde{\Phi}_n(t) = 
\min_{\substack{S \subseteq \Csig[\Xbf] \\ \widetilde\pi(S) \leq t}} 
\Phi(S; G_{n,r}[\Csig[X]]),
$$
where \smash{$\widetilde\pi$} is the stationary distribution over
$G_{n,r}[\Csig[\Xbf]]$. The utility of the conductance function comes from the
known upper bound of mixing time by  
\begin{equation}
\label{eqn: average_conductance}
\int \frac{1}{t \widetilde{\Phi}_n(t)^2} dt,
\end{equation}
which results in a tighter bound than merely using the conductance when
\smash{$\widetilde{\Phi}_n(t)$} is large for small values of $t$.

Our proof relies on a novel (to the best of our knowledge) uniform lower bound
on this conductance function over $G_{n,r}$ in terms of a population-level
analogue, which we denote \smash{$\widetilde{\Phi}_{\Pbb,r}$}. Plugging
\smash{$\widetilde{\Phi}_{\Pbb,r}$} into \eqref{eqn: average_conductance} yields
a mixing time bound (with respect to total variation distance) on the order $d 
D^2/r^2 + d^2\log(D/r)$ over convex sets. By contrast, \eqref{eqn:
  inverse_mixing_time_lower_bound} is of order $d^2 D^2/r^2 + d^3\log(D/r)$ 
(ignoring the $o_r(1)$ term) but handles mixing time with respect to relative
pointwise distance (which is known to be a stricter metric). 
\end{remark}

@@

\begin{remark}
	For convenience and maximum generality, we define $\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)$ to correspond with the bound given by \eqref{eqn: inverse_mixing_time_lower_bound_nonconvex}, and assume only \ref{asmp: cluster_separation} - \ref{asmp: low_noise_density}. However, if we additionally have \ref{asmp: convex}, then we could sharpen $\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)$ to the tighter rate of \eqref{eqn: inverse_mixing_time_lower_bound}, with nothing changing hereafter. 
\end{remark}

>>>>

\begin{abstract}
\vskip 0.1 in % RESOLVED
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.
\end{abstract}

@@

When applied to geometric graphs constructed from a large number of samples,
global spectral clustering methods can be computationally cumbersome and   
insensitive to the local geometry of the underlying distribution
\citep{leskovec2010,mahoney2012}.  This has led to recent increased interest in
local spectral algorithms, which leverage locally-biased spectra computed using
random walks around a user-specified seed node.  A popular local clustering
algorithm is Personalized PageRank (PPR), first introduced by
\citet{haveliwala2003}, and further developed by
\citet{spielman2011,spielman2014,andersen2006,mahoney2012,zhu2013},
among others.

@@

Here we define a pair of criteria that reflect the quality of a cluster with
respect to $G=(V,E,w)$.  There are many graph-based measures of cluster quality  
that one could consider; see, e.g., \citet{yang2015,fortunato2010} for 
an overview.  The pair of criteria that we focus on are (arguably) quite
natural, and moreover, they play a fundamental role in our analysis 
of the PPR algorithm.  Our two criteria capture the \emph{external} and
\emph{internal} connectivity of a subset $S \subseteq V$, denoted $\Phi(S; G)$
and $\Psi(S; G)$, respectively, and defined below in turn.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the rotation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationary distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ m: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}},  
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
% AJG: RESOLVED
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

@@

For a level $\beta > 0$ and a target stationary measure $\pi_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: p_u > \beta \pi_{0}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{3}{10},\frac{1}{2})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{3}{10}, \frac{1}{2})$} is standard in the analysis of PPR
algorithms, see, e.g., \citet{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary measure $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{3}{10}, \frac{1}{2})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\widehat{C} = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{3}{10}, \frac{1}{2})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

\item In Section \ref{sec: experiments}, we empirically
  demonstrate that violations of the geometric conditions we require manifestly
  impact density cluster recovery. 
\end{enumerate}

@@

\subsection{Related Work}
%In addition to the background given above, a few related lines of work are worth
%highlighting. For neighborhood graphs of the type we consider, continuous analogues to (for
%instance) normalized cut have been defined, over the data-manifold rather than
%the graph, and convergence finite sample graph-theoretic functionals to their
% continuous counterparts has been shown 
% \cite{garciatrillos16, arias-castro12, maier11}.

@@

Building on earlier work of
\citet{koltchinskii2000}, \citet{vonluxburg2008,hein2005} studied the limiting behaviour of spectral clustering
algorithms. These authors show that when samples are obtained from a
distribution, and we appropriately construct a geometric graph, the spectrum of
the Laplacian converges to that of the Laplace-Beltrami operator on the
data-manifold. However, relating the partition obtained using the
Laplace-Beltrami operator to the more intuitively defined high-density
clusters can be challenging in general.

@@

\item 
\label{asmp: low_noise_density}
\emph{Low noise density:} There exists $\gamma > 0$ such that for all $x
\in \Rd$ with $0 < \dist(x, \Csig) \leq \sigma$,   
$$
  \inf_{x' \in \Csig} f(x') - f(x) \geq  \dist(x, \Csig)^{\gamma},
$$
where \smash{$\dist(x,\Aset) = \inf_{x_0 \in \Aset} \norm{x - x_0}$}.
% RJT: should we change this to c_0?  c_1 seems to be used below and throughout
% for other constants, and c_0 is not used

@@

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} along with \ref{asmp:  
cluster_separation} are similar to the $(\sigma,\epsilon)$-saliency of
\citet{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citet{singh2009} (as well as
many other works on density clustering and level set estimation.) It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as PPR.

In the next several subsections, we will derive bounds on the cluster quality
criteria evaluated on ($\sigma$-expansions of) density clusters. For notational simplicity,
hereafter for $S \subseteq V$, we will abbreviate $\Phi(S; G_{n,r})$ by
$\Phi_{n,r}(S)$, and similarly, $\Psi(S; G_{n,r})$ by $\Psi_{n,r}(S)$, and
$\tau_{\infty}(G_{n,r}[S])$ by $\tau_{n,r}(S)$. We will also use $\nu$ for
Lebesgue measure on $\Rd$, and $\nu_d = \nu(B)$ for the measure of the unit ball
$B=B(0,1)$.  
% RJT: you had B(0,d) before ... just check this should have been B(0,1) 
% AJG: Yep.

@@

\begin{theorem}
\label{thm: conductance_upper_bound}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density}, and
\ref{asmp: cluster_separation}, for some 
$\sigma, \lambda_{\sigma}, c_1, \gamma > 0$. Moreover, let
\begin{equation}
\label{eqn: weighted_cluster_volume}
\nu_{\Pbb}(\Csig) \leq \frac{1}{2}
\end{equation}
where $\nu_{\Pbb}(\Aset) = \int_{\Aset} f(x) dx$ for $\Aset \subset \Rd$.
 Then for any $0 < r <
\sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and
\begin{equation}
\label{eqn: conductance_sample_complexity}
n \geq \frac{9\log(2/\delta)}{\epsilon^2}\left(\frac{1}
  { \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
\end{equation}
we have
\begin{equation}
  \label{eqn: conductance_additive_error_bound}
  \frac{\Phi_{n,r}(\Csig[\mathbf{X}])}{r} \leq 4 c_{\sigma} d
  \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
    \frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
\end{equation}
with probability at least $1-\delta$, where $c_{\sigma} = 1 / \sigma$. 
\end{theorem}

@@

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is almost
tight. Specifically, choosing $\Asig = B(0,\sigma)$ and
$$
 f(x)  = \begin{cases}
   \lambda &\text{for $x \in \Asig$}, \\
   \lambda - \dist(x,\Asig)^{\gamma} & \text{ for $0 < \dist(x,\Asig) < r$}, 
\end{cases}
$$
we have that for $n$ on the order of the lower bound in \eqref{eqn: 
  conductance_sample_complexity}, 
$$
  \frac{\Phi_{n,r}(\Asig[\mathbf{X}])}{r} \geq c \frac{(\lambda -
    \frac{r^{\gamma+1}}{\gamma+1})}{\lambda} - \epsilon,
$$
with probability at least $1 - \delta$, for some constant $c$. (Note that the
factor of $1 / \sigma$ in $c_{\sigma}$ is not replicated above.)
\end{remark}

@@

Next we lower bound the inverse mixing time \eqref{eqn: inv_mixing_time} of 
$\Csig[\Xbf]$, or equivalently, as $\Psi_{n,r}(\Csig[\Xbf]) =
1/\tau_{n,r}(\Csig[\Xbf])$, we upper bound the mixing time. 
% RJT: note I changed the next two theorems to be upper bounds on the mixing
% time.  Upper bounds on the inverse inverse mixing time, which was a bit
% superfluous.  Just make sure that the supplement is in sync with this
% AJG: Perfect.

\begin{theorem}
\label{thm: inverse_mixing_time_lower_bound_nonconvex}
Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
Assumptions \ref{asmp: bounded_density} and \ref{asmp: cluster_diameter} 
for some $\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D > 0$. Then for any $0 <
r < \sigma/(4d)$, $0 < \delta < 1$, $\epsilon > 0$, and $n$ satisfying 
$$
\sqrt{3^{d+1}\frac{(\log n + d\log \mu + \log(4/\delta))}{n \nu_d r^d
    \lambda_{\sigma}}} \leq \epsilon, 
$$
where \smash{$\mu = \log(\frac{2D}{r})$}, we have
\begin{multline}
\label{eqn: inverse_mixing_time_lower_bound_nonconvex}
\tau_{n,r}(\Csig[\Xbf]) \leq (d \log \mu + c_{\lambda}) \cdot \\  
\left(c_1 + 9^d c_2 \frac{\Lambda_{\sigma}^4D^{2d}}{\lambda_{\sigma}^4r^{2d}}\left(d
    \log \mu + c_{\lambda}\right)\right),
\end{multline}
with probability at least $1 - \delta$, where $c_1,c_2>0$ are universal constants
 and $c_{\lambda} = \log(\Lambda_{\sigma}^2/ 
\lambda_{\sigma}^2)$.   
\end{theorem}
% RJT: saying that c_1,c_2 depend on the dimension is a little odd/weak.
% c_2 multiplyies, e.g., a factor of d \log \mu.  So if it's d^{10}, then this
% will swamp the linear factor of ....
% AJG: Fixed.

\begin{remark}
The proof of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} relies
on upper bounding the mixing time using the \emph{conductance} of $G_{n,r}[\Csig[\Xbf]]$, 
$$
\widetilde{\Phi} = \min_{S \subseteq \Csig[\Xbf]} \Phi(S; G_{n,r}[\Csig[\Xbf]]).
$$
The factor of $1/r^{2d}$ in the bound in \eqref{eqn:
  inverse_mixing_time_lower_bound_nonconvex} is suboptimal. This exponential 
dependence on $d$ stems from a loose bound on the aforementioned
conductance. In particular, we assert only that any set $S \subseteq \Csig[X]$
must have $\cut(S; \Csig[\Xbf])$ on the order of $n^2 r^{2d}$, while upper bounding
$\vol(S; \Csig[\Xbf])$ by roughly $n^2 r^d$, for a bound on the conductance of
order $r^d$. The presence of $r^{2d}$ comes from upper bounding the mixing time
by about \smash{$1/\widetilde{\Phi}^2$}, this being a variant of classic results
on rapid mixing \citep{jerrum89}.  
\end{remark}

@@

\begin{remark}
The only potentially exponential dependence on dimension comes from the
factor of $c_d$. However, for sufficiently small values of $r$ $o_r(1)/r^2$ will be dominated by $1/r^2$, and therefore the last term in \eqref{eqn: inverse_mixing_time_lower_bound} will contribute negligibly to the overall bound.
\end{remark}
\begin{remark}
We achieve superior rates in Theorem \ref{thm: inverse_mixing_time_lower_bound}
to those of Theorem \ref{thm: inverse_mixing_time_lower_bound_nonconvex} in part 
by working with a generalization of the conductance, the \emph{conductance
  function},
$$
\widetilde{\Phi}_{n,r}(t) = 
\min_{\substack{S \subseteq \Csig[\Xbf] \\ \widetilde\pi(S) \leq t}} 
\Phi(S; G_{n,r}[\Csig[X]]),
$$
where \smash{$\widetilde\pi$} is the stationary distribution over
$G_{n,r}[\Csig[\Xbf]]$. The utility of the conductance function comes from the
known upper bound of mixing time by  
\begin{equation}
\label{eqn: average_conductance}
\int \frac{1}{t \widetilde{\Phi}_n(t)^2} dt,
\end{equation}
which results in a tighter bound than merely using the conductance when
\smash{$\widetilde{\Phi}_n(t)$} is large for small values of $t$.

Our proof relies on a novel (to the best of our knowledge) lower bound
on this conductance function over $G_{n,r}$ in terms of a population-level
analogue, which we denote \smash{$\widetilde{\Phi}_{\Pbb,r}$}, and define formally in the supplementary document.
\begin{lemma}
	\label{lem: consistency_of_conductance_function_main}
	Fix $0 < t < 1/2$. Under the conditions on $\Csig$ given by Theorem \ref{thm: inverse_mixing_time_lower_bound}, the following statement holds: with probability one, as $n \to \infty$,
	\begin{equation}
	\label{eqn: pointwise_consistency_of_conductance_function}
	\liminf_{n \to \infty} \widetilde{\Phi}_{n,r}(t) \geq \min\left\{\widetilde{\Phi}_{\Pbb,r}(t), c_d r \omega_r(1)\right\}
	\end{equation}
	where $c_d$ is the same as in Theorem \ref{thm: inverse_mixing_time_lower_bound}.
\end{lemma}
Plugging
\smash{$\widetilde{\Phi}_{\Pbb,r}$} into \eqref{eqn: average_conductance} yields
a mixing time bound (with respect to total variation distance) on the order $d 
D^2/r^2 + d^2\log(D/r)$ over convex sets. By contrast, \eqref{eqn:
  inverse_mixing_time_lower_bound} is of order $d^2 D^2/r^2 + d^3\log(D/r)$ 
(ignoring the $o_r(1)$ term) but handles mixing time with respect to relative
pointwise distance (which is known to be a stricter metric). 
\end{remark}

@@

\begin{remark}
	For convenience and maximum generality, we define $\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)$ to correspond with the bound given by \eqref{eqn: inverse_mixing_time_lower_bound_nonconvex}, and assume only \ref{asmp: bounded_density} - \ref{asmp: cluster_diameter}. However, if we additionally have \ref{asmp: convex}, then we could sharpen $\mathbf{\Psi}(\sigma, \lambda_{\sigma}, \Lambda_{\sigma}, D)$ to the tighter rate of \eqref{eqn: inverse_mixing_time_lower_bound}, with no other changes to subsequent results. 
\end{remark}
